Word Sense Disambiguation\\ \small{Leveraging the NLTK Toolkit and External Tools}

### 2011-04-14 paper/paper.tex

@@

We  chose to implement our system in Python, and chose to use the machine
learning algorithms in NLTK which is also Python-based.  We also used NLTK for its part of speech tagger, dependecy parser, and to import the Senseval data.  We coded the rest of the assignment on our own.

@@

We used a \ward stepwise approach to indentifying optimal feature combinations.
We first ran the system with all features on, producing a total of $256=2^8$
combinations.  We then used backwards stepwise regression to identify feature
combinations with a small number of features but with high performance

@@

The NLTK was used to implement two machine learning algorithms, naive Bayes and
decision trees.

@@

The NLTK implements a naïve bayes classifier which chooses the best sense given a
dictionary of features and assigns confidence probabilities to the chosen
sense. Naive bayes does assume an independence of features. The probability of
each feature contributes to the overall choice.

@@

NLTK provides probability measures for each of its guesses when using the
maximum entropy and the naive Bayes learning methods.  We harnessed these to
determine the cutoff for making a guess, or resorting to a ``U'' if we were
unconfident in the tagging.  Occasional unknowns also occur because such
unknowns are present in the training data, and we made no effort to remove
them, so the learning algorithm views them as any other label.

@@

We also used nltk's part of speech tagger to tag the context, and included those
as we did the literal words in the feature set.  This was trained on nltk's
\verb+maximum entropy treebank pos tagger+ data, which is included with the nltk
package.

\subsection{Cooccurrence}

Looks through the training data and compiles a list of most common words for
each item in the data. When the test data is run, the feature looks for the
created vector and uses that instead of trying to make a new one. Next the
extractor takes a window of words around the head word and generates the 0, 1
tuple of features for inclusion to the dictionary. This word list is used as the
keys of the feature vector when using the test data as a list of words that may
cooccur with each item in the data. The value associated with each key in the
feature vector starts out as 0 (for occurring 0 times in the window near a test
word) If a key in the feature vector does occur with a word in the test data,
the value associated with the key changes to 1 even if the key occurs more than
once.  This is necessary because the machine learning algorithm we used does not
attempt to analyze the keys beyond simple equality, so higher numbers than one
could have masked the possibly more important distinction between 0 and not-0.
In other words, keeping track of the instances of occurrence vs clustering them
into 2 groups reduces the sparsity of the data and allows for a higher chance of
accurate classifications based on the limited training data that we have. This
is an especially important consideration since the feature vectors that select
the most common words by looking through all the instances of unique lexical
items in the training data are unsupervised.

@@

\begin{itemize}
\item Classifier (naive Bayes or decision list)
\item Collocation window length (0--8)
\item Cooccurrence handling (on or off)
\item Base word handling (on or off)
\item Bootstrap iterations (0--4)
\item Dependency parsing (on or off)
\end{itemize}

The predictor combination yeilding the lowest value of the Akaike information criterion
included the following parameters.

\begin{itemize}
\item Intercept
\item Classifier
\item Collocation window size
\item Coocurrence
\item Base word handling
\item Classifier$\times$collocation window size interaction
\item Classifier$\times$cooccurrence interaction
\end{itemize}

@@

The naive Bayes classifier had better top-range performance.

@@

Our run with the highest overall F-measure (overall precision 0.718, overall
recall 0.610), used cooccurrence, colocation out to a distance of 2, base word,
no dependency parsing, and bootstrapped once:

@@

An idea for a feature extractor was to take the most common words in a context
of a lexeme and use those words as features. Based off of the cooccurrence code,
this would generate a vector from the training data, find the most common words
from the test data, and then run the collected data through the classifier.

Unfortunately, this feature extractor did not give very good results. Both the
precision and recall for this feature extractor alone on the EnglishLS test data
were 0.039. This is to be expected since it is essentially the cooccurrence
feature extractor but without grouping data into buckets. The sparsity problem
is even more noticeable in this extractor.

@@

Collocation seemed to be a fairly powerful disambugator, which is to be
expected, since function words near to the target word can be very strong
predictors of how the word is being used in context.  Effectiveness was
strongest with words close to the target, and weaker the farther away we
extended the feature.

>>>>

\newcommand{\naive}{na\"ive}
\newcommand{\Naive}{Na\"ive}

@@

We  chose to implement our system in Python, and chose to use the machine
learning algorithms in NLTK which is also Python-based.  We also used NLTK for
its part of speech tagger, dependency parser, and to import the Senseval data.
We coded the rest of the assignment on our own.

@@

We used a \ward stepwise approach to identifying optimal feature combinations.
We first ran the system with all features on, producing a total of $256=2^8$
combinations.  We then used backwards stepwise regression to identify feature
combinations with a small number of features but with high performance

@@

The NLTK was used to implement two machine learning algorithms, \naive{} Bayes and
decision trees.

@@

The NLTK implements a \naive{} Bayes classifier which chooses the best sense given a
dictionary of features and assigns confidence probabilities to the chosen
sense. \Naive{} Bayes does assume an independence of features. The probability of
each feature contributes to the overall choice.

@@

NLTK provides probability measures for each of its guesses when using the
maximum entropy and the \naive{} Bayes learning methods.  We harnessed these to
determine the cutoff for making a guess, or resorting to a ``U'' if we were not
confident in the tagging.  Occasional unknowns also occur because such unknowns
are present in the training data, and we made no effort to remove them, so the
learning algorithm views them as any other label.

@@

We also used NLTK's part of speech tagger to tag the context, and included those
as we did the literal words in the feature set.  This was trained on NLTK's
\verb+maximum entropy treebank pos tagger+ data, which is included with the NLTK
package.

\subsection{Co-occurrence}

Looks through the training data and compiles a list of most common words for
each item in the data. When the test data is run, the feature looks for the
created vector and uses that instead of trying to make a new one. Next the
extractor takes a window of words around the head word and generates the 0, 1
tuple of features for inclusion to the dictionary. This word list is used as the
keys of the feature vector when using the test data as a list of words that may
co-occur with each item in the data. The value associated with each key in the
feature vector starts out as 0 (for occurring 0 times in the window near a test
word) If a key in the feature vector does occur with a word in the test data,
the value associated with the key changes to 1 even if the key occurs more than
once.  This is necessary because the machine learning algorithm we used does not
attempt to analyze the keys beyond simple equality, so higher numbers than one
could have masked the possibly more important distinction between 0 and not-0.
In other words, keeping track of the instances of occurrence vs clustering them
into 2 groups reduces the sparsity of the data and allows for a higher chance of
accurate classifications based on the limited training data that we have. This
is an especially important consideration since the feature vectors that select
the most common words by looking through all the instances of unique lexical
items in the training data are unsupervised.

@@

\begin{itemize}
\item Classifier (\naive{} Bayes or decision list)
\item Collocation window length (0--8)
\item Co-occurrence handling (on or off)
\item Base word handling (on or off)
\item Bootstrap iterations (0--4)
\item Dependency parsing (on or off)
\end{itemize}

The predictor combination yielding the lowest value of the Akaike information criterion
included the following parameters.

\begin{itemize}
\item Intercept
\item Classifier
\item Collocation window size
\item Co-occurrence
\item Base word handling
\item Classifier$\times$collocation window size interaction
\item Classifier$\times$co-occurrence interaction
\end{itemize}

@@

The \naive{} Bayes classifier had better top-range performance.

@@

Our run with the highest overall F-measure (overall precision 0.718, overall
recall 0.610), used co-occurrence, collocation out to a distance of 2, base word,
no dependency parsing, and bootstrapped once:

@@

An idea for a feature extractor was to take the most common words in a context
of a lexeme and use those words as features. Based off of the co-occurrence code,
this would generate a vector from the training data, find the most common words
from the test data, and then run the collected data through the classifier.

Unfortunately, this feature extractor did not give very good results. Both the
precision and recall for this feature extractor alone on the EnglishLS test data
were 0.039. This is to be expected since it is essentially the co-occurrence
feature extractor but without grouping data into buckets. The sparsity problem
is even more noticeable in this extractor.

@@

Collocation seemed to be fairly powerful at disambiguating, which is to be
expected, since function words near to the target word can be very strong
predictors of how the word is being used in context.  Effectiveness was
strongest with words close to the target, and weaker the farther away we
extended the feature.
